\chapter{Fermionic Field Eigenstates, Coherent States, and Spin--Statistics}

Consider the Dirac Hamiltonian operator \cite{haag1996local}
\begin{equation}
\hat H
=
\int d^d x \,
\hat\psi^\dagger(x)
\left(
-i\bm{\alpha}\cdot\nabla + \beta m
\right)
\hat\psi(x).
\end{equation}

One may attempt to define field eigenstates
\begin{equation}
\hat\psi(x)\ket{\psi}
=
\psi(x)\ket{\psi},
\end{equation}
analogously to bosonic field eigenstates.

At first sight, it seems plausible that
\begin{equation}
\bra{\psi'}\hat H\ket{\psi}
\stackrel{?}{=}
H_{\mathrm{cl}}[\psi]\,
\delta[\psi'-\psi].
\end{equation}

However, this intuition fails for fermions.

The Dirac field satisfies anticommutation relations:
\begin{equation}
\{
\hat\psi(x),
\hat\psi(y)
\}
=
0.
\end{equation}

Suppose
\begin{equation}
\hat\psi(x)\ket{\psi}
=
\psi(x)\ket{\psi}.
\end{equation}

Then
\begin{align}
\hat\psi(x)\hat\psi(y)\ket{\psi}
&=
\psi(x)\psi(y)\ket{\psi},
\\
\hat\psi(y)\hat\psi(x)\ket{\psi}
&=
\psi(y)\psi(x)\ket{\psi}.
\end{align}

Since the operators anticommute,
\begin{equation}
\hat\psi(x)\hat\psi(y)
=
-
\hat\psi(y)\hat\psi(x),
\end{equation}
consistency requires
\begin{equation}
\psi(x)\psi(y)
=
-
\psi(y)\psi(x).
\end{equation}

Therefore the eigenvalues themselves must anticommute. This forces the introduction of Grassmann variables.

However, there is a deeper obstruction. Anticommuting operators cannot be simultaneously diagonalized. If all operators \(\hat\psi(x)\) were diagonal in some basis, they would effectively act as multiplication operators, which commute. Therefore a genuine fermionic field basis does not exist in the Hilbert space.

This can already be seen in the simplest example of a single fermionic mode:
\begin{equation}
\{
\hat a,
\hat a^\dagger
\}
=
1.
\end{equation}

In the occupation number basis,
\begin{equation}
\hat a
=
\begin{pmatrix}
0 & 1 \\
0 & 0
\end{pmatrix}.
\end{equation}

The annihilation operator is nilpotent:
\begin{equation}
\hat a^2 = 0,
\end{equation}
and is not diagonalizable. It possesses only the eigenvalue zero and admits no complete basis of eigenvectors inside the Hilbert space.

Thus the object \(\ket{\psi}\) cannot belong to the physical Hilbert space in the ordinary sense.

Instead, fermions are described using coherent states.

For bosons, coherent states satisfy
\begin{equation}
\hat a\ket{\alpha}
=
\alpha\ket{\alpha},
\qquad
\alpha\in\mathbb C.
\end{equation}

For fermions, coherent states satisfy
\begin{equation}
\hat\psi(x)\ket{\eta}
=
\eta(x)\ket{\eta},
\end{equation}
where \(\eta(x)\) is Grassmann-valued.

They are constructed as
\begin{equation}
\ket{\eta}
=
\exp\left(
\int d^d x\,
\eta(x)\hat\psi^\dagger(x)
\right)\ket{0}.
\end{equation}

For a single fermionic mode,
\begin{equation}
\ket{\eta}
=
\ket{0}
+
\eta\ket{1}.
\end{equation}

These coherent states are not ordinary Hilbert-space vectors with complex coefficients. Rather, they live in an extended algebraic object,
\begin{equation}
\mathcal H \otimes \Lambda,
\end{equation}
where \(\Lambda\) denotes the Grassmann algebra.

The overlap of coherent states is
\begin{equation}
\braket{\eta'}{\eta}
=
\exp\left(
\int d^d x\,
\eta'^*(x)\eta(x)
\right).
\end{equation}

The resolution of the identity becomes
\begin{equation}
\mathbf 1
=
\int
\mathcal D\eta^*
\mathcal D\eta\,
e^{-\int \eta^*\eta}
\ket{\eta}\bra{\eta}.
\end{equation}

The matrix element of the Hamiltonian between coherent states takes the form
\begin{equation}
\bra{\eta'}\hat H\ket{\eta}
=
H_{\mathrm{cl}}[\eta'^*,\eta]
\exp\left(
\int d^d x\,
\eta'^*(x)\eta(x)
\right).
\end{equation}

This replaces the bosonic structure
\begin{equation}
\bra{\phi'}\hat H\ket{\phi}
\sim
H[\phi]\delta[\phi'-\phi].
\end{equation}

The Hilbert space itself remains an ordinary complex vector space. The fermionic Fock space is
\begin{equation}
\mathcal H
=
\bigoplus_n
\wedge^n \mathcal H_1,
\end{equation}
and states possess ordinary complex coefficients.

Grassmann variables are not elements of the Hilbert space. Rather, they provide an auxiliary algebraic coordinate system adapted to fermionic anticommutation relations.

The deeper origin of these structures lies in the spin--statistics connection.

In relativistic quantum field theory, locality and Lorentz invariance force half-integer spin fields to satisfy anticommutation relations,
\begin{equation}
\{
\hat\psi(x),
\hat\psi(y)
\}
=
0,
\end{equation}
while integer-spin fields satisfy commutation relations.

Because fermionic fields anticommute, they cannot admit a simultaneous spectral decomposition analogous to bosonic field operators. Consequently, there is no fermionic analog of a classical configuration-space basis.

Berezin calculus and Grassmann coherent states replace ordinary spectral theory in the fermionic case.

\section*{Electromagnetic Neutrality of Neutrinos}

In the electroweak theory, electric charge is given by
\begin{equation}
Q
=
T_3
+
\frac{Y}{2},
\end{equation}
where \(T_3\) denotes weak isospin and \(Y\) denotes hypercharge.

The left-handed lepton doublet is
\begin{equation}
\begin{pmatrix}
\nu_L \\
e_L
\end{pmatrix},
\end{equation}
with hypercharge
\begin{equation}
Y=-1.
\end{equation}

For the neutrino,
\begin{equation}
T_3 = +\frac12,
\end{equation}
and therefore
\begin{equation}
Q
=
+\frac12
-
\frac12
=
0.
\end{equation}

For the electron,
\begin{equation}
T_3 = -\frac12,
\end{equation}
so
\begin{equation}
Q
=
-\frac12
-
\frac12
=
-1.
\end{equation}

Thus neutrinos are electrically neutral because they belong to the trivial representation of the electromagnetic gauge group \(U(1)_{\mathrm{em}}\).

They do not couple directly to the electromagnetic field because their electric charge vanishes. However, they still interact through the weak interaction via the \(W^\pm\) and \(Z^0\) gauge bosons.

Therefore, neutrinos do not lack a \(U(1)\) symmetry. Rather, they transform trivially under the electromagnetic \(U(1)\).