
\chapter{Superselection sectors}

\chapterquote{...it must be equally possible to interpret the
data on the assumption that the charged $\pi$-meson field
is scalar, provided corresponding modifications in the
properties of the other fields are made. }{Wick, Wightman, Wigner, 1952}

The following discussion is mostly based in Wick, 
Wightman and Wigner's original work \cite{Wick1952sectors} where they 
introduced superselection sectors, the idea being that rather than 
states being defined as rays in the Hilbert space they are equivalent
up to relative phases between certain partitions of the Hilbert space. 
The more modern algebraic approach is taken from Diego Pontello's thesis \cite{pontello2019entanglement}. 

\section{A pedestrian point of view}
The idea of superselection sectors is quite simple. 
Suppose that you have a quantum mechanical system with Hilbert space
$\mathcal{H}=\bigoplus_i \mathcal{H}_i$, such that any state 
$\ket{\psi}$ decomposes uniquely,
\begin{equation*}
    \ket{\psi}=\sum_{i=1}^{n}\ket{\psi,i}, \; \ket{\psi,i} \in \mathcal{H}_i.
\end{equation*}
Now, the claim is that:
\begin{center}
\itshape
\parbox{0.75\textwidth}{
\centering
$\mathcal{H}_i$ are \textbf{superselection sectors} if there is no
\textbf{physical} measurement one can do to tell the state
$\ket{\psi}=\sum_{i=1}^{n} \ket{\psi,i}$ apart from
$\ket{\psi'}=\sum_{i=1}^{n} \omega_i \ket{\psi,i}$,
where $\omega_i$ are phases.\par
}
\end{center}

An equivalent way of rephrasing this is that,

\begin{center}
\itshape
\parbox{0.75\textwidth}{
    \centering
    Any observable $O$ with a non-null matrix element
    $\bra{\psi,i} O \ket{\psi,j}$ cannot be realized
    physically\footnotemark.
}
\end{center}

\footnotetext{
This is the "super" of superselection, as it is a stronger
condition than the usual selection rules
$\bra{\varphi}\hat H\ket{\psi}=0$.
}

As the expectation value $\bra{\psi'} O \ket{\psi'}$ would allow 
us to know something about the phases $\omega_i$. This means we have
to restrict our selves to observables $\mathcal{O}=\bigoplus_{i=1}^{n} O_i$ that are 
diagonal by blocks in $\mathcal{H}_i$. This has a remarkable consequence:
if the only experiments we can do are those that don't mix sectors,
then any coherent superposition of different sectors is indistinguishable
from a mixed state, as
\begin{align*}
\bra{\psi'}\mathcal{O}\ket{\psi'} &= \sum_{i=1}^{n}\bra{\psi,i}O_i\ket{\psi,i}\\
                                  &= \mathrm{tr}(\rho \mathcal{O}).
\end{align*}
With $\rho = \sum_{i=1}^{n}\bra{\psi,i}\ket{\psi,i}\ket{\psi,i}\bra{\psi,i}$\footnote{
    You can check this by extending $\ket{\psi,i}$ to an orthonormal
in each subspace $\mathcal{H}_i$.}.
We have only talked about observables i.e. hermitian operators. Is natural to 
think (or at least should be after the next section) that any operation that
mixes sectors - like a unitary - should be forbidden physically. This restricts
the algebra of operators available in a laboratory to be block diagonal 
$\bigoplus_{i=1}^n \mathcal{B}(\mathcal{H}_i)$\footnote{
    $\mathcal{B}(\mathcal{H})$ is notation for the operators that act on 
    $\mathcal{H}$ or the bounded operators if you are thinking 
    infinite dimension.
}. 

\subsection{Superselection sectors and symmetry}
So far all seems too ad-hoc, we arbitrarily decomposed the Hilbert space in
orthogonal sectors and claimed that you can't tell relative phases between 
the states in each one. However, this happens naturally if you consider
that you can't tell the difference between states related by a symmetry 
operation.

Take for instance time reversal symmetry $\mathcal{T}= U\mathcal{K}$,
$U$ the unitary part and $\mathcal{K}$ the complex conjugation. The statement
that time reversed partners are indistinguishable is the equallytie 
of the two probabilities\footnotemark for any observable $\mathcal{O}$,

\footnotetext{As Winger pointed out, the requirement for symmetries is that they
keep probabilities invariant rather than amplitudes, allowing antiunitaries.}

\begin{equation*}
    \abs{\bra{\varphi}\mathcal{O}\ket{\psi}}^2
    =
    \abs{\bra{\mathcal{T}\varphi}\mathcal{O}\ket{\mathcal{T}\psi}}^2
\end{equation*}
Which is valid if we choose operators that satisfy,
\begin{equation*}
    \mathcal{K}(U^\dagger\mathcal{O}U) = \mathcal{O}.
\end{equation*}
Now take a fermionic state  $\ket{f}$ and a bosonic one $\ket{b}$,
for fermions $\mathcal{T}^2=-1$ and for bosons $\mathcal{T}^2=1$. The matrix
element of $\mathcal{O}$ then reads,
\begin{align}
    \begin{split}
    \bra{f}\mathcal{O}\ket{b}&
    =\bra{f}\mathcal{K}(U^\dagger\mathcal{O}U)\ket{b},\\
    &=\overline{\bra{\mathcal{K}f}U^\dagger\mathcal{O}U\ket{\mathcal{K}b}},\\
    &=\overline{\bra{\mathcal{T}f}\mathcal{O}\ket{\mathcal{T}b}}.
    \label{time_reversal_1}
    \end{split}
\end{align}
Now one more time,

\begin{equation}
    \overline{\bra{\mathcal{T}f}\mathcal{O}\ket{\mathcal{T}b}}
    =
    \overline{\overline{\bra{\mathcal{T}^2f}\mathcal{O}\ket{\mathcal{T}^2b}}}
    =-\bra{f}\mathcal{O}\ket{b}.
    \label{time_reversal_2}
\end{equation}
And from \ref{time_reversal_1} and \ref{time_reversal_2} we see that 
 $\bra{f}\mathcal{O}\ket{b}=0$, so any operation that respects time reversal 
 symmetry gets neatly separated into a bosonic and fermionic sector!
 \subsection{Charge}
\section{The algebraic point of view}
In the algebraic point of view one starts from an observable algebra 
$\mathfrak{A}$, this is an abstract algebra, which means is not necessarily 
a sub algebra of $\mathcal{B}(\mathcal{H})$. However, one can take a morphism 
between algebras $\pi: \mathfrak{A} \rightarrow \mathcal{B}(\mathcal{H})$ 
which defines a \textbf{representation} of the algebra $\mathfrak{A}$ in 
the Hilbert space $\mathcal{H}$. As it happens with groups,
this  representation might be reducible\footnotemark. If the $\pi$
admits a decomposition in irreps\footnotemark, then naturally the operators
$\pi(a)$ decompose in sectors blocks. Of course one could choose an 
irreducible representation sticking, to a single sector.

But what happens with states? The whole point of using abstract algebras
is to write statements independent of representations. For an abstract
algebra the states form a positive and normalized convex subset of the dual
space $\mathcal{S}(\mathfrak{A}) \subset \mathfrak{A}^*$. In two words: states
are functionals that map elements of the algebra to expectation values
and pure states live in the corners.  


\footnotetext{As with groups, $\pi$ is reducible if there exist a proper 
subspace $\mathcal{S}\subset \mathcal{H}$ such that 
$\pi(a)(\mathcal{S})\subset\mathcal{S}$ 
for all elements $a$ of $\mathfrak{A}$.}
\footnotetext{Sadly, for algebras, is often not the case.}
\section{Why don't we learn about SS in QM?}

\section{Final words}

In the quantum mechanics point of view one chooses the superselection
sector based on sensible physical assumption -i.e. you can't superpose a 
fermion with a boson-, in the algebraic approach it seems that the sectors
where already given to us by the algebra. Yet, an election of superselection
sectors has still been made, it's hidden in which algebra we use to model
physically realizable operations.   


\begin{center}
\textbf{Acknowledgments}

Thanks to Leandro Martinek for the fruitful discussions that leaded
to these notes. 
\end{center}

